Detection of Treatment-Subgroup Interactions in Clustered Datasets with Generalized Linear Mixed-effects Model Trees (GLMM tree)

### 2015-08-18 pkg/glmertree/paper/treatment_interactions paper.tex
git-svn-id: svn://svn.r-forge.r-project.org/svnroot/partykit@748 255f4249-2263-4417-b1ea-8498779e27a7
git-svn-id: svn://svn.r-forge.r-project.org/svnroot/partykit@749 255f4249-2263-4417-b1ea-8498779e27a7

\usetikzlibrary{shapes,shadows,arrows,positioning}
\title{Detection of Treatment-Subgroup Interactions in Clustered Datasets with Generalized Linear Mixed-effects Model (GLMM) Trees}
\shorttitle{GLMM Trees}
\leftheader{GLMM Trees}
\rightheader{GLMM Trees}
\author{M. Fokkema$^1$, N. Smits$^2$, A. Zeileis$^3$, T. Hothorn$^4$, H. Kelderman$^5$}
\affiliation{$^1$Universiteit Leiden, $^2$Universiteit van Amsterdam, $^3$Universit\"{a}t Innsbruck, $^4$Universit\"{a}t Z\"{u}rich, $^5$Universiteit Leiden and Vrije Universiteit, Amsterdam}

%\abstract{}

@@

\section{Abstract}
Identification of subgroups of patients for which treatment A is more effective than treatment B, and vice versa, is of key importance to the development of personalized medicine. Several tree-based algorithms have been developed for the detection of such treatment-subgroup interactions. In many instances, however, datasets may have a clustered structure, where observations are clustered within, for example, research centers, studies or persons. In the current paper we propose a new algorithm, GLMM tree, that allows for detection of treatment-subgroup interactions, as well as estimation of cluster-specific random effects. The algorithm uses model-based recursive partitioning (MOB) to detect treatment-subgroup interactions, and a generalized linear mixed-effects model for estimation of random-effects parameters. In a simulation study, we evaluate the performance of GLMM tree and compare it with that of MOB trees without random-effects estimation. In datasets without treatment-subgroup interactions, GLMM tree was found to have a much lower Type I error rate than MOB trees without random effects (4 and 33\%, respectively). Furthermore, in datasets with treatment-subgroup interactions, GLMM tree recovered the true treatment subgroups much more often than MOB without random effects (90\% and 61\% of the datasets, respectively). Also, GLMM tree predicted treatment outcome differences more accurately than MOB trees without random effects (average accuracy of .94 and .88, respectively). We illustrate the application of GLMM tree on a patient-level dataset of a meta-analysis on the effects of psycho- and pharmacotherapy for depression. We conclude that GLMM tree is a promising algorithm for the detection of treatment-subgroup interactions in clustered datasets, and discuss some directions for future research.

\textit{Keywords}: model-based recursive partitioning, treatment-subgroup interactions, random effects, generalized linear mixed-effects model, classification and regression trees

%\section{Citation stuff for journal publication}
%
%Potential Psychological Methods references: only \citeA{StroyMall09} (not added yet). Maybe force in: 
%
%''Do logistic regression and signal detection identify different subgroups at risk? Implications for the design of tailored interventions. Kiernan, Michaela; Kraemer, Helena C.; Winkleby, Marilyn A.; King, Abby C.; Taylor, C. Barr. Psychological Methods, Vol 6(1). ''
%
%''Hedges, L. V., \& Pigott, T. D. (2004). The power of statistical tests for moderators in meta-analysis. Psychological methods, 9(4), 426.''
%
%''Cooper, H., \& Patall, E. A. (2009). The relative benefits of meta-analysis conducted with individual participant data versus aggregated data. Psychological methods, 14(2), 165.'' (they write in abstract ''Regardless of whether a meta-analysis is conducted with IPD or AD, synthesists must remain vigilant in how they interpret their results. They must avoid ecological fallacies, Simpson’s paradox, and interpretation of synthesis-generated evidence as supporting causal inferences.'')
%
%''Hedges, L. V., \& Vevea, J. L. (1998). Fixed-and random-effects models in meta-analysis. Psychological methods, 3(4), 486.''
%
%Statistics in Medicine references: \cite{DussyMech14,FostyTayl11,HiggyWhit01,LipkyDmit11}

@@

Tree-based methods are such hypothesis-generating methods, as they can automatically detect subgroups which differ on the expected outcomes for one or more treatments. Due to their flexibility, tree-based methods are preeminently suited to the detection of treatment-subgroup interactions: they can handle many potential predictor variables at once and can automatically detect (higher order) interactions between predictor variables. Several promising tree-based algorithms for the detection of treatment-subgroup interactions have been developed (e.g., \citeNP{DussyMech14, DussyMeul04, SuyTsai09, FostyTayl11, LipkyDmit11, ZeilyHoth08}; see \citeNP{DoovyDuss14} for an overview). Among these methods, model-based recursive partitioning (MOB; \citeNP{ZeilyHoth08}) seems to be the most flexible tool for detecting treatment-subgroup interactions, as it offers a very generic data-analytic framework for detecting partitions in a dataset, with different model parameter estimates. The recursive partitioning in MOB can be based on a broad class of parametric models that can be fitted using M-type estimators \cite{ZeilyHoth08}, the most well-known example being the generalized linear model (GLM). Earlier, GLM-based MOB has been successfully applied by \citeA{DrieySmit14} in the detection of subgroups with differential treatment outcomes for two different psychotherapies.

However, none of the aforementioned tree-based algorithms allow for taking into account the clustered structure of datasets. In many cases, researchers may want to detect treatment-subgroup interactions in datasets with a clustered structure \cite<e.g., >{KoopyHeij07}. For example, in individual-level patient data meta-analyses, in which datasets of multiple trials evaluating the effects of the same treatments are pooled. In such analyses, the clustered structure of the dataset should be taken into account by including study-specific effects in the model, prompting the need for modeling random effects \cite<e.g., >{Frie93,DersyLair86,HiggyWhit01}. Likewise, longitudinal datasets, and datasets from multi-center trials also require modeling of random effects. Ignoring the clustered structure of datasets may lead to biased inference, due to underestimated standard errors \cite<e.g., >{BrykyRaud92,Hox98,NooryOpde05}. More specifically, when the interest is in subgroup detection, ignoring random effects on the outcome variable may result in the detection of spurious subgroups \cite<e.g., >{SelaySimo12}.

@@

\subsection{GLM}
In a clinical trial, where the outcomes of two or more treatments are compared, an overall GLM may be used to estimate treatment effects. The goal is to estimate a model for predicting the value of a treatment outcome $Y$, which follows, for example, a normal, binomial or Poisson distribution \footnote{ an overview of notation used is provided in the Appendix}:

\begin{equation}
\label{eq:expected value}
	E[y_i | x_i] = \mu_i 
\end{equation}

@@

To keep notation and examples simple, we assume $x_{i}^{\top}$ and $\beta$ to have length 2. That is, the effects of only two treatment conditions are estimated and no additional covariates are included in the GLM. However, additional treatment conditions and covariates can easily be included.

As an example, the estimated GLM of the artificial motivating dataset is graphically represented in Figure \ref{fig:fixedeffects}. The boxplots in Figure \ref{fig:fixedeffects} show the distribution of the posttreatment depression scores in both treatment groups. There seems to be little overall difference in effects of both treatments, as the slope of the regression line is nearly zero. We shall see that this does not necessarily mean that posttreatment depression score and treatment type are unrelated, as the effect of treatment may be moderated by variables not yet included in the model.

@@

\section{Appendix: Notation}

>>>>

\usetikzlibrary{shapes,shadows,arrows,positioning}
\title{Detection of Treatment-Subgroup Interactions in Clustered Datasets with Generalized Linear Mixed-effects Model Trees (GLMM tree)}
\shorttitle{GLMM Tree}
\leftheader{GLMM Tree}
\rightheader{GLMM Tree}
\author{M. Fokkema$^1$, N. Smits$^2$, A. Zeileis$^3$, T. Hothorn$^4$, H. Kelderman$^5$}
\affiliation{$^1$Universiteit Leiden, $^2$Universiteit van Amsterdam, $^3$Universit\"{a}t Innsbruck, $^4$Universit\"{a}t Z\"{u}rich, $^5$Universiteit Leiden and Vrije Universiteit, Amsterdam}

\acknowledgements{The authors would like to thank Prof. Pim Cuijpers, Prof. Jeanne Miranda, Dr. Boadie Dunlop, Prof. Rob DeRubeis, Prof. Zindel Segal, Dr. Sona Dimidjian, Prof. Steve Hollon and Erica Weitz for granting access to the dataset for the application. The work for this paper was partially done while MF, AZ and TH were visiting the Institute for Mathematical Sciences, National University of Singapore in 2013. The visit was supported by the Institute.}

@@

\section{Abstract}
Identification of subgroups of patients for which treatment A is more effective than treatment B, and vice versa, is of key importance to the development of personalized medicine. Several tree-based algorithms have been developed for the detection of such treatment-subgroup interactions. In many instances, however, datasets may have a clustered structure, where observations are clustered within, for example, research centers, studies or persons. In the current paper we propose a new algorithm, GLMM tree, that allows for detection of treatment-subgroup interactions, as well as estimation of cluster-specific random effects. The algorithm uses model-based recursive partitioning (MOB) to detect treatment-subgroup interactions, and a generalized linear mixed-effects model for estimation of random-effects parameters. In a simulation study, we evaluate the performance of GLMM tree and compare it with that of MOB trees without random-effects estimation. In datasets without treatment-subgroup interactions, GLMM tree was found to have a much lower Type I error rate than MOB trees without random effects (4 and 33\%, respectively). Furthermore, in datasets with treatment-subgroup interactions, GLMM tree recovered the true treatment subgroups much more often than MOB without random effects (90\% and 61\% of the datasets, respectively). Also, GLMM tree predicted treatment outcome differences more accurately than MOB trees without random effects (average accuracy of .94 and .88, respectively). We illustrate the application of GLMM tree on a patient-level dataset of a meta-analysis on the effects of psycho- and pharmacotherapy for depression. We conclude that GLMM tree is a promising algorithm for the detection of treatment-subgroup interactions in clustered datasets, and discuss some directions for future research.\\
\\
\textit{Keywords}: model-based recursive partitioning, treatment-subgroup interactions, random effects, generalized linear mixed-effects model, classification and regression trees

%\section{Citation stuff for journal publication}
%
%Potential Psychological Methods references: only \citeA{StroyMall09} (not added yet). Maybe force in: 
%
%''Hedges, L. V., \& Pigott, T. D. (2004). The power of statistical tests for moderators in meta-analysis. 
%
%''Cooper, H., \& Patall, E. A. (2009). The relative benefits of meta-analysis conducted with individual participant data versus aggregated data. Psychological methods, 14(2), 165.'' (they write in abstract ''Regardless of whether a meta-analysis is conducted with IPD or AD, synthesists must remain vigilant in how they interpret their results. They must avoid ecological fallacies, Simpson’s paradox, and interpretation of synthesis-generated evidence as supporting causal inferences.'')
%
%''Hedges, L. V., \& Vevea, J. L. (1998). Fixed-and random-effects models in meta-analysis. Psychological methods, 3(4), 486.''

@@

Tree-based methods are such hypothesis-generating methods, as they can automatically detect subgroups which differ on the expected outcomes for one or more treatments. Due to their flexibility, tree-based methods are preeminently suited to the detection of treatment-subgroup interactions: they can handle many potential predictor variables at once and can automatically detect (higher order) interactions between predictor variables \cite{StroyMall09}. Several promising tree-based algorithms for the detection of treatment-subgroup interactions have been developed (e.g., \citeNP{DussyMech14, DussyMeul04, SuyTsai09, FostyTayl11, LipkyDmit11, ZeilyHoth08}; see \citeNP{DoovyDuss14} for an overview). Among these methods, model-based recursive partitioning (MOB; \citeNP{ZeilyHoth08}) seems to be the most flexible tool for detecting treatment-subgroup interactions, as it offers a very generic data-analytic framework for detecting partitions in a dataset, with different model parameter estimates. The recursive partitioning in MOB can be based on a broad class of parametric models that can be fitted using M-type estimators \cite{ZeilyHoth08}, the most well-known example being the generalized linear model (GLM). Earlier, GLM-based MOB has been successfully applied by \citeA{DrieySmit14} in the detection of subgroups with differential treatment outcomes for two different psychotherapies.

However, none of the aforementioned tree-based algorithms allow for taking into account the clustered structure of datasets. In many cases, researchers may want to detect treatment-subgroup interactions in datasets with a clustered structure \cite<e.g., >{KoopyHeij07}. For example, in individual-level patient data meta-analyses, in which datasets of multiple trials evaluating the effects of the same treatments are pooled. In such analyses, the clustered structure of the dataset should be taken into account by including study-specific effects in the model, prompting the need for modeling random effects \cite<e.g., >{CoopyPatt09,HiggyWhit01}. Likewise, longitudinal datasets, and datasets from multi-center trials also require modeling of random effects. Ignoring the clustered structure of datasets may lead to biased inference, due to underestimated standard errors \cite<e.g., >{BrykyRaud92,NooryOpde05}. More specifically, when the interest is in subgroup detection, ignoring random effects on the outcome variable may result in the detection of spurious subgroups \cite<e.g., >{SelaySimo12}.

@@

\subsection{GLM}
In a clinical trial, where the outcomes of two or more treatments are compared, an overall GLM may be used to estimate treatment effects. The goal is to estimate a model for predicting the value of a treatment outcome, which follows, for example, a normal, binomial or Poisson distribution \footnote{ an overview of notation used is provided in the Appendix}:

\begin{equation}
\label{eq:expected_value}
	E[y_i | x_i] = \mu_i 
\end{equation}

@@

To keep notation and examples simple, we assume $x_{i}^{\top}$ and $\beta$ to have length 2. That is, the effects of only two treatment conditions are estimated and no additional covariates are included in the GLM. However, additional treatment conditions and covariates can easily be included. In addition, examples and datasets in the current paper will focus on continuous response variables with normally distributed errors, such as posttreatment severity of a disorder. But the models and algorithms to be discussed can also be applied with discrete outcomes, such as remission of a disorder (yes/no).

To illustrate, the GLM estimated for the artificial motivating dataset is graphically represented in Figure \ref{fig:fixedeffects}. The boxplots in Figure \ref{fig:fixedeffects} show the distribution of the posttreatment depression scores in both treatment groups. There seems to be little overall difference in effects of both treatments, as the slope of the regression line is nearly zero. We shall see that this does not necessarily mean that posttreatment depression score and treatment type are unrelated, as the effect of treatment may be moderated by variables not yet included in the model.

@@

\newpage
\section{Appendix: Notation}
